Information Aggregation in Exponential Family Markets

### 2014-04-11 WriteUp/maxentpredmkts.tex
more changes including comments in response file

%new commands
\newcommand{\la}{\langle}
\newcommand{\ra}{\rangle}
\newcommand{\mX}{\mathcal{X}}
\newcommand{\bR}{\mathbf{R}}
\newcommand{\grad}{\nabla}
\newcommand{\Exp}{\mathbf{E}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\q}{\mathbf{q}}
\newcommand{\mM}{\ensuremath{\mathcal{M}}}
\newcommand{\hmu}{\ensuremath{\hat{\mu}}}
\newcommand{\mP}{\ensuremath{\mathcal{P}}}
\newcommand{\mY}{\ensuremath{\mathcal{Y}}}
\newcommand{\mS}{\ensuremath{\mathcal{S}}}
\newcommand{\ds}{\displaystyle}
\newcommand{\mF}{\mathcal{F}}
\newcommand{\mD}{\mathcal{D}}
\newcommand{\htheta}{\hat{\theta}}
\newcommand{\ttheta}{\tilde{\theta}}
%---------------------------------------------------------------------------
\newtheorem{thm}{Theorem}[section]
\newtheorem{prop}[thm]{Proposition}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{con}[thm]{Conjecture}
\newtheorem{cor}[thm]{Corollary}
\newtheorem{ex}[thm]{Example}
\newtheorem{qn}[thm]{Question}
\newcommand{\defn}{\bigbreak\noindent{\bf Definition. }}
\newcommand{\undefn}{\bigbreak}
\newcommand{\rmk}{\bigbreak\noindent{\bf Remark. }}
\newcommand{\unrmk}{\bigbreak}
%\newcommand{\proof}{\noindent{\bf Proof. }}
\newcommand{\unproof}{\hfill$\Box$\bigbreak}
\newcommand{\defeq}{\stackrel{\mathrm{def}}{=}}
\newcommand{\holds}[1]{\stackrel{?}{#1}}
\newcommand{\conv}{\mathrm{conv}}
\newcommand{\ie} {{\it i.e. }}
\newcommand{\eg} {{\it e.g. }}
\newcommand{\etal} {{\it et al. }}
\newcommand{\E}{\mathbf{E}}
\newcommand{\betavec}{\pmb{\beta}}
\newcommand{\qvec}{\mathbf{q}}
\newcommand{\lpf}{\mathbf{\psi}} %logpartition function
\newcommand{\family}{\mathcal F} 
\newcommand{\suff}{\pmb{\phi}} % suff stats
\newcommand{\muvec}{\pmb{\mu}}
\newcommand{\st}{\hspace{10pt}\mathrm{s.t.}\hspace{10pt}}
\newcommand{\norm}[1]{|| #1 ||}
\newcommand{\tp}{\tilde{p}}
\newcommand{\CE}{C\!E}

@@

% \jake{I want to add more to this list. It's ok but doesn't totally reflect what's in the last few sections.}

@@

@@

The following observation is crucial. Let $\tp \in \mP$ be a density (not necessarily from an exponential family) with expected statistic $\mu$, let $p(\cdot\,;\mu)$ be the exponential family with the same expected statistic, and let $\hmu \in \mM$ be an alternative report. Then from~(\ref{exp-fam}) note
%
\begin{equation} \label{equalizer-rule}
\Exp_{\tp}[\log p(x;\hmu)] = \Exp_{p(\cdot;\mu)}[\log p(x;\hmu)] = \la \htheta, \mu \ra - T(\htheta),
\end{equation}
%
where $\htheta = \grad G(\hmu)$ is the natural parameter for the exponential family with statistic $\hmu$. We see from this that the expected log score only depends on the expectation $\mu$ of the underlying density, not the full density, which is how we can achieve proper scoring according to Definition~{\ref{def:proper-score}.
%
\begin{theorem} \label{maxent-score}
Consider the logarithmic scoring rule $S(\mu,x) = \log p(x;\mu)$ defined over a set of densities $\mD$ parametrized by $\mM$. The scoring rule is proper if and only if $\mD$ is the exponential family with statistic $\phi$. 
\end{theorem}
%
%
\begin{proof}
Let $\mu,\hmu \in \mM$ be the agent's true belief and an alternative report, and let $p \in \mP$ be a density consistent with $\mu$. Let $\theta = \grad G(\mu)$ and $\htheta = \grad G(\hmu)$, and note that $\mu = \grad T(\theta)$. We have
%
\begin{eqnarray}
\Exp_p[\log p(x;\mu)] - \Exp_p[\log p(x;\hmu)]
& = & \la \theta, \mu \ra - T(\theta) - \la \htheta, \mu \ra + T(\htheta) \nonumber\\
& = & T(\htheta) - T(\theta) - \la \htheta - \theta, \mu \ra \nonumber\\
& = & T(\htheta) - T(\theta) - \la \htheta - \theta, \grad T(\theta) \ra. \label{breg-div}
\end{eqnarray}
%
The latter is positive by the strict convexity of $T$, which shows that the log score is proper. For the converse, assume the defined log score is proper. By the Savage characterization of proper scoring rules for expectations (see~\citet[Thm.\ 1]{gneiting2007strictly} and~\citet{Savage71}), we must have
%
$$
S(\mu,x) = G(\mu) - \la \grad G(\mu), \mu - \phi(x) \ra
$$
for some strictly convex function $G$. Let $T = G^*$, so that $\grad G = {\grad T}^{-1}$, and let $\theta = \grad G(\mu)$. Then the above can be written as
%
\begin{eqnarray*}
\log p(x;\mu) & = & G(\mu) - \la \grad G(\mu), \mu - \phi(x) \ra \\
& = & \la \theta, \mu \ra - T(\theta) - \la \theta, \mu - \phi(x) \ra = \la \theta, \phi(x) \ra - T(\theta),
\end{eqnarray*}
%
which shows that $p(x;\mu)$ takes the form of an exponential family.
\end{proof}
%

@@

An advantage of this generalization of the log score is that, for many domains (multi-dimensional included) and expectations of interest, it leads to novel closed-form scoring rules. By examining the log densities of various exponential families, we can for instance obtain scoring rules for several different combinations of the arithmetic, geometric, and harmonic means, as well as higher order moments. The following examples illustrate the construction.
%
\begin{example} \label{ex:exponential}
As base measure we take the Lebesgue restricted to $[0,+\infty)$, and we consider the statistic $\phi(x) = x$ so that we are simply eliciting the mean. The maximum entropy distribution with a given mean $\mu$ is the exponential distribution, and taking its log density gives the scoring rule
%
\begin{equation} \label{exp-scoring}
S(\mu, x) = -\frac{x}{\mu} - \log \mu.
\end{equation}
%
We stress that although this rule is derived from the exponential distribution, Theorem~\ref{maxent-score} implies that it elicits the mean of any distribution supported on the non-negative reals (e.g., Pareto, lognormal). Indeed, it is easy to see that the expected score~(\ref{exp-scoring}) depends only on the mean of the agent's belief because it is linear in $x$. 
%
%Typically the exponential distribution is parametrized by its rate $\lambda = 1/\mu$. In that case the rule takes the form $S(\mu, x) = -\lambda x + \log \lambda$, and we see that $\lambda$ corresponds to the natural parameter.
%
As a generalization of this example, the maximum entropy distribution for the $k$-th moment $\phi(x) = x^k$ with respect to the same base measure is the Weibull distribution. Taking its log density leads to the following equivalent scoring rule:
%
\begin{equation} \label{weibull-scoring}
S(\mu, x) = (k-1) \log x - k \log \mu - \Gamma\left(1+\frac{1}{k}\right)^k \left(\frac{x}{\mu}\right)^k,
\end{equation}
%
where $\Gamma$ denotes the gamma function (the extension of the factorial to the reals). We have not found either scoring rule~(\ref{exp-scoring}) or~(\ref{weibull-scoring}) in the literature.
\end{example}
%
%
\begin{example} \label{ex:gaussian}
As a base measure we take the Lebesgue over the real numbers $\bR$. We are interested in eliciting the mean $\mu$ and variance $\sigma^2$, so as a statistic we take $\phi(x) = (x, x^2)$ for which $\Exp_p[\phi(x)] = (\mu, \mu^2+\sigma^2)$. The max entropy distribution for a given mean and variance is the Gaussian, whose log density gives the scoring rule
%
\begin{equation}
S((\mu, \sigma^2), x) = -\frac{(x-\mu)^2}{\sigma^2} - \log \sigma^2.
\end{equation}
%
Again, we stress that this scoring rule elicits the mean and variance of any density over the real numbers, not just those of a normal distribution. The construction easily generalizes to a multi-dimensional outcome space by taking the log density of the multivariate normal:
%
\begin{equation}
S((\mu, \Sigma), x) = -(x - \mu)'\Sigma^{-1}(x - \mu) - \log |\Sigma|.
\end{equation}
%
Here the statistics being elicited are the mean vector $\mu$ and the covariance matrix $\Sigma$. These scoring rules have been studied by~\citet{dawid1999coherent} as rules that only depend on the mean and variance of the reported density. They note that these rules are weakly proper (because they do not distinguish between densities with the same first and second moments), but do not make the point that knowledge of the full density is not necessary on the part of the agent.
\end{example}
%

@@

In this section, we consider the evolution of the market state when traders are budget-limited.
% and their budgets are known to the market maker. We consider a cost-function based market with cost function $C$ and payoff $\phi(x)$ with non-negative payoff. 
We assume that the traders trade in multiple instances of the market. That is, after the initial trades, the outcome is revealed and the traders are paid off. Then, the market runs again with the same set of traders and the outcome drawn from a (possibly different) exponential family distribution. The goal of this section is to analyze how the budget of traders evolve under these circumstances. As before, the market price is interpreted as a probability density over the outcome space and the share vector as the natural parameter of an exponential family distribution. Consistent with the connections drawn in Section~\ref{sec:scoring} and throughout, we measure the error in prediction using the standard log loss.

@@

\paragraph{Limited Damage}
We will now quantify the error in prediction that the market maker might have to endure as a result of ill-informed entities entering the market. We assume that these entities trade in multiple instances of the market; thus the exposure of the market maker is over several rounds. The log loss function for $\theta$ shares held is defined as $L(\theta,x)=-\log p(x;\theta)=C(\theta)-\la \theta\phi(x)\ra$.

>>>>

%new commands
\newcommand{\la}{\langle}
\newcommand{\ra}{\rangle}
\newcommand{\mX}{\mathcal{X}}
\newcommand{\bR}{\mathbf{R}}
\newcommand{\grad}{\nabla}
\newcommand{\Exp}{\mathbf{E}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\q}{\mathbf{q}}
\newcommand{\mM}{\ensuremath{\mathcal{M}}}
\newcommand{\hmu}{\ensuremath{\hat{\mu}}}
\newcommand{\hsigma}{\ensuremath{\hat{\sigma}}}
\newcommand{\hSigma}{\ensuremath{\hat{\Sigma}}}
\newcommand{\mP}{\ensuremath{\mathcal{P}}}
\newcommand{\mY}{\ensuremath{\mathcal{Y}}}
\newcommand{\mS}{\ensuremath{\mathcal{S}}}
\newcommand{\ds}{\displaystyle}
\newcommand{\mF}{\mathcal{F}}
\newcommand{\mD}{\mathcal{D}}
\newcommand{\htheta}{\hat{\theta}}
\newcommand{\ttheta}{\tilde{\theta}}
%---------------------------------------------------------------------------
\newtheorem{thm}{Theorem}[section]
\newtheorem{prop}[thm]{Proposition}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{con}[thm]{Conjecture}
\newtheorem{cor}[thm]{Corollary}
\newtheorem{ex}[thm]{Example}
\newtheorem{qn}[thm]{Question}
\newcommand{\defn}{\bigbreak\noindent{\bf Definition. }}
\newcommand{\undefn}{\bigbreak}
\newcommand{\rmk}{\bigbreak\noindent{\bf Remark. }}
\newcommand{\unrmk}{\bigbreak}
%\newcommand{\proof}{\noindent{\bf Proof. }}
\newcommand{\unproof}{\hfill$\Box$\bigbreak}
\newcommand{\defeq}{\stackrel{\mathrm{def}}{=}}
\newcommand{\holds}[1]{\stackrel{?}{#1}}
\newcommand{\conv}{\mathrm{conv}}
\newcommand{\ie} {{\it i.e. }}
\newcommand{\eg} {{\it e.g. }}
\newcommand{\etal} {{\it et al. }}
\newcommand{\E}{\mathbf{E}}
\newcommand{\betavec}{\pmb{\beta}}
\newcommand{\qvec}{\mathbf{q}}
\newcommand{\lpf}{\mathbf{\psi}} %logpartition function
\newcommand{\family}{\mathcal F} 
\newcommand{\suff}{\pmb{\phi}} % suff stats
\newcommand{\muvec}{\pmb{\mu}}
\newcommand{\st}{\hspace{10pt}\mathrm{s.t.}\hspace{10pt}}
\newcommand{\norm}[1]{|| #1 ||}
\newcommand{\tp}{\tilde{p}}
\newcommand{\CE}{C\!E}

@@

 A note about play-money markets: Even in prediction markets set up with play-money, risk aversion is plausible since it captures the fact that a trader high on the leaderboard may not want to risk her standing with large late trades. Further, in such markets especially, budget constraints are key since they help alleviate the problem that traders can easily manipulate prices with unreasonable trades. Our results are therefore particularly relevant in markets with play-money.
% \jake{I want to add more to this list. It's ok but doesn't totally reflect what's in the last few sections.}

@@

Some commonly studied distributions that are in fact exponential families include the binomial, beta and normal distributions. To see for instance, that the normal distribution is an exponential family note that instead of the usual parametrization using mean  $\mu$ and  variance $\sigma$, it can be parametrized in terms of $\vec{\theta}=(\theta_{1},\theta_{2})$ where $\theta_{1}=\frac{\mu}{\sigma^{2}}$ and $\theta_{2}=\frac{-1}{2\sigma^{2}}$,  with corresponding log partition function $-\frac{\beta_{1}^{2}}{4\beta_{2}}-\frac{1}{2}\log(-2\beta_{2})$.

@@

The following observation is crucial. Let $\tp \in \mP$ be a density (not necessarily from an exponential family) with expected statistic $\mu$, let $p(\cdot\,;\mu)$ be the exponential family with the same expected statistic, and let $\hmu \in \mM$ be an alternative report. Then from~(\ref{exp-fam}) note
%
\begin{equation} \label{equalizer-rule}
\Exp_{\tp}[\log p(x;\hmu)] = \Exp_{p(\cdot;\mu)}[\log p(x;\hmu)] = \la \htheta, \mu \ra - T(\htheta),
\end{equation}
%
where $\htheta = \grad G(\hmu)$ is the natural parameter for the exponential family with statistic $\hmu$. Note that the expected log score does not depend  on the entire maximum entropy distribution but just its expected statistic $\mu$. This is how we can achieve proper scoring according to Definition~{\ref{def:proper-score}.
%We see from this that the expected log score only depends on the expectation $\mu$ of the underlying density, not the full density, which is how we can achieve proper scoring according to Definition~{\ref{def:proper-score}.
%
%
\begin{theorem} \label{maxent-score}
Consider the logarithmic scoring rule $S(\mu,x) = \log p(x;\mu)$ defined over a set of densities $\mD$ parametrized by $\mM$. The scoring rule is proper if and only if $\mD$ is the exponential family with statistic $\phi$, where $\mu=\E_{p(x;\mu)}[\phi(x)]$. 
\end{theorem}
%
%
\begin{proof}
Let $\mu,\hmu \in \mM$ be the agent's true belief and an alternative report, and let $p \in \mP$ be a density consistent with $\mu$. Let $\theta = \grad G(\mu)$ and $\htheta = \grad G(\hmu)$, and note that $\mu = \grad T(\theta)$. We have
%
\begin{eqnarray}
\Exp_p[\log p(x;\mu)] - \Exp_p[\log p(x;\hmu)]
& = & \la \theta, \mu \ra - T(\theta) - \la \htheta, \mu \ra + T(\htheta) \nonumber\\
& = & T(\htheta) - T(\theta) - \la \htheta - \theta, \mu \ra \nonumber\\
& = & T(\htheta) - T(\theta) - \la \htheta - \theta, \grad T(\theta) \ra. \label{breg-div}
\end{eqnarray}
%
The latter is positive by the strict convexity of $T$, which shows that the log score is proper. For the converse, assume the defined log score is proper. By the Savage characterization of proper scoring rules for expectations (see~\citet[Thm.\ 1]{gneiting2007strictly} and~\citet{Savage71}), we must have
%
$$
S(\mu,x) = G(\mu) - \la \grad G(\mu), \mu - \phi(x) \ra
$$
for some strictly convex function $G$. Let $T = G^*$, so that $\grad G = {\grad T}^{-1}$, and let $\theta = \grad G(\mu)$. Then the above can be written as
%
\begin{eqnarray*}
\log p(x;\mu) & = & G(\mu) - \la \grad G(\mu), \mu - \phi(x) \ra \\
& = & \la \theta, \mu \ra - T(\theta) - \la \theta, \mu - \phi(x) \ra = \la \theta, \phi(x) \ra - T(\theta),
\end{eqnarray*}
%
which shows that $p(x;\mu)$ takes the form of an exponential family.
\end{proof}
%

@@

An advantage of this generalization of the log score is that, for many domains (multi-dimensional included) and expectations of interest, it leads to novel closed-form scoring rules. By examining the log densities of various exponential families, we can for instance obtain scoring rules for several different combinations of the arithmetic, geometric, and harmonic means, as well as higher order moments. The following examples illustrate the construction. Here, we have used $\mu$ and $\sigma$ for the true mean and variance and $\hmu$ and $\hsigma$ for the elicited mean and variance respectively (and $\hSigma$ for the variance of the multivariate normal).
%
\begin{example} \label{ex:exponential}
As base measure we take the Lebesgue restricted to $[0,+\infty)$, and we consider the statistic $\phi(x) = x$ so that we are simply eliciting the mean. The maximum entropy distribution with a given mean $\mu$ is the exponential distribution, and taking its log density gives the scoring rule
%
\begin{equation} \label{exp-scoring}
S(\hmu, x) = -\frac{x}{\hmu} - \log \hmu.
\end{equation}
%
We stress that although this rule is derived from the exponential distribution, Theorem~\ref{maxent-score} implies that it elicits the mean of any distribution supported on the non-negative reals (e.g., Pareto, lognormal). Indeed, it is easy to see that the expected score~(\ref{exp-scoring}) depends only on the mean of the agent's belief because it is linear in $x$. 
%
%Typically the exponential distribution is parametrized by its rate $\lambda = 1/\mu$. In that case the rule takes the form $S(\mu, x) = -\lambda x + \log \lambda$, and we see that $\lambda$ corresponds to the natural parameter.
%
As a generalization of this example, the maximum entropy distribution for the $k$-th moment $\phi(x) = x^k$ with respect to the same base measure is the Weibull distribution. Taking its log density leads to the following equivalent scoring rule:
%
\begin{equation} \label{weibull-scoring}
S(\hmu, x) = (k-1) \log x - k \log \hmu - \Gamma\left(1+\frac{1}{k}\right)^k \left(\frac{x}{\hmu}\right)^k,
\end{equation}
%
where $\Gamma$ denotes the gamma function (the extension of the factorial to the reals). We have not found either scoring rule~(\ref{exp-scoring}) or~(\ref{weibull-scoring}) in the literature.
\end{example}
%
%
\begin{example} \label{ex:gaussian}
As a base measure we take the Lebesgue over the real numbers $\bR$. We are interested in eliciting the mean $\mu$ and variance $\sigma^2$, so as a statistic we take $\phi(x) = (x, x^2)$ for which $\Exp_p[\phi(x)] = (\mu, \mu^2+\sigma^2)$. The max entropy distribution for a given mean and variance is the Gaussian, whose log density gives the scoring rule
%
\begin{equation}
S((\hmu, \hsigma^2), x) = -\frac{(x-\hmu)^2}{\hsigma^2} - \log \hsigma^2.
\end{equation}
%
Again, we stress that this scoring rule elicits the mean and variance of any density over the real numbers, not just those of a normal distribution. The construction easily generalizes to a multi-dimensional outcome space by taking the log density of the multivariate normal:
%
\begin{equation}
S((\hmu, \hSigma), x) = -(x - \hmu)'\hSigma^{-1}(x - \hmu) - \log |\hSigma|.
\end{equation}
%
Here the statistics being elicited are the mean vector $\mu$ and the covariance matrix $\Sigma$. These scoring rules have been studied by~\citet{dawid1999coherent} as rules that only depend on the mean and variance of the reported density. They note that these rules are weakly proper (because they do not distinguish between densities with the same first and second moments), but do not make the point that knowledge of the full density is not necessary on the part of the agent.
\end{example}
%

@@

In this section, we consider the evolution of the market state when traders are budget-limited.
% and their budgets are known to the market maker. We consider a cost-function based market with cost function $C$ and payoff $\phi(x)$ with non-negative payoff. 
We assume that the traders trade in multiple instances of the market. That is, after the initial setup and trades, the outcome is revealed and the traders are paid off. Then, the market runs again with the same set of traders and the outcome drawn from a (possibly different) exponential family distribution. The goal of this section is to analyze how the budget of traders evolve under these circumstances. As before, the market price is interpreted as a probability density over the outcome space and the share vector as the natural parameter of an exponential family distribution. Consistent with the connections drawn in Section~\ref{sec:scoring} and throughout, we measure the error in prediction using the standard log loss.

@@

Note that while an uninformative trader is unconstrained in his belief, the informative trader's belief corresponds to the parameter of the true distribution.
\paragraph{Limited Damage}
We will now quantify the error in prediction that the market maker might have to endure as a result of ill-informed entities entering the market. We assume that these entities trade in multiple instances of the market; thus the exposure of the market maker is over several rounds. The log loss function for $\theta$ shares held is defined as $L(\theta,x)=-\log p(x;\theta)=C(\theta)-\la \theta\phi(x)\ra$.
